\documentclass[11pt]{article}

\usepackage{a4wide}
\usepackage{xspace}
\usepackage[draft]{fixme}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}

\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}

\newcommand{\VAL}{\textsf{VAL}\xspace}
\newcommand{\CAL}{\textsf{CAL}\xspace}
\newcommand{\THEO}{\textsf{THEO}\xspace}
\newcommand{\ATOM}{\textsf{ATOM}\xspace}
\newcommand{\NOM}{\textsf{NOM}\xspace}
\newcommand{\FORM}{\textsf{FORM}\xspace}

\begin{document}

\title{Hybrid Type Theory\\ A Quartet in Four Movements}
\author{Carlos Areces \\
\and
Patrick Blackburn
\and
Antonia Huertas
\and 
Mara Manzano
}
\date{}
\maketitle

\begin{abstract}
This paper is about a logic\ldots and beauty. It tells the story of an idea that 
was born by putting together the work of four great names in in the history of 
Logic. 

The ideas of Reichenbach about the importance of temporal reference and
Prior about the importance of the internal perspective provided by modal and
tense logics have already been combined in~\cite{Blackburn1994}, In this
paper, simple hybrid logic, named propositional nominal tense logic, allowed
to put Reichenbach and Prior together. A decade later, in~\cite{ArecesBlackburn2005}, nominals were added to Montague's Intensional Logic to build
a richer hybrid logic (nominal intensional logic). The resulting system was
capable of capturing the insights of Reichenbach, Prior and Montague together.

What we shall do now is to add Leon Henkin's ideas of completeness proofs in
higher order systems to nominal intensional logic, which is presented here
as a hybrid type theory. Thus, this paper is about a very expressive formal
language, that is, type theory enhanced with hybrid modal resources. 
%Our
%collaboration with Carlos Areces and Patrick Blackburn grew 
%from~\cite{ArecesBlackburn2005}, producing the definition of a formal calculus for
%hybrid type theory and its completeness proof. 
As it is such a strong logic,
we looked towards Henkin's use of non-standard models in higher order logics
and we were able to adapt Henkin's general models thereby following in his
footsteps, and being loyal not only to his conception but also to the
construction Henkin proposed in his 1950 paper~\cite{Henkin1950}. Thus we
added Leon Henkin to the list of maestros that inspired this work.
We are sure Newton Da Costa will enjoy the quartet.
\end{abstract}

\begin{small}
\tableofcontents
\end{small}

\input{introduction.tex}
\input{modallogic.tex}
\input{typetheory.tex}
\input{conclusions.tex}

\bibliographystyle{plain} 
\bibliography{dacosta10}

\end{document}
